---
title: "Machine Learning HW 1"
author: "Jonah Bonesteel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in data and libraries

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(ROCit)
library(earth)
library(mgcv)

df <- read_csv("insurance_t.csv")
#View(df)

```

Missing value imputation

```{r}

#data_types <- sapply(df, class)
#print(data_types)

# Ordinal variables
      
ordinal_df = data.frame(df$DDA,df$DIRDEP,df$NSF,df$SAV,df$ATM,df$CD,df$IRA,
                        df$INV,df$MM,df$MMCRED,df$CC,df$CCPURC,df$SDB,df$INAREA, df$INS)

Mode <- function(x) {
  uniq_x <- unique(x)
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}

ordinal_df <- ordinal_df %>% 
  mutate_all(~ ifelse(is.na(.), Mode(.), .))
ordinal_df <- data.frame(lapply(ordinal_df, factor))
# Numeric variables

cont_df = data.frame(df$ACCTAGE,df$DDABAL,df$DEP,df$DEPAMT,df$CHECKS,df$NSFAMT,df$PHONE,df$TELLER,df$SAVBAL,
                        df$ATMAMT,df$POS,df$POSAMT,df$CDBAL,df$IRABAL,df$INVBAL,df$MMBAL,
                        df$CCBAL,df$INCOME,df$LORES,df$HMVAL,df$AGE,df$CRSCORE)

cont_df <- cont_df %>%
  mutate_all(~ ifelse(is.na(.), median(., na.rm = TRUE), .))

imp_df <- cbind(ordinal_df,cont_df)

remove_prefix <- function(data, prefix) {
  names(data) <- sub(paste0("^", prefix), "", names(data))
  return(data)
}

# Remove the 'df.' part at the beginning of the column names:
imp_df <- remove_prefix(imp_df, "df.")

```

Create a flagged dataframe

```{r}
binary_df <- data.frame(sapply(df, function(x) as.numeric(is.na(x))))
binary_df <- data.frame(lapply(binary_df, as.factor))
# Add "_flag" to the end of each column name in the binary dataframe
colnames(binary_df) <- paste0(colnames(binary_df), "_flag")

final_df <- cbind(imp_df, binary_df)
# drop flagged columns where there are no imputed (flagged) values
zero_columns <- colSums(final_df == 0) == nrow(final_df)
final_df <- final_df[, !zero_columns]

```

MARS model 1 (no cross-validation):

```{r}

mars1 <- earth(INS ~ ., data = final_df, glm = list(family=binomial))
summary(mars1)

```

Predict and test

```{r}
test_mars1_df <- final_df

test_mars1_df$phat_mars1 <- predict(mars1, type = 'response')

roc_mars1 <- rocit(as.vector(test_mars1_df$phat_mars1), test_mars1_df$INS)

summary(roc_mars1)

```

AUC = 0.799

GAM Model using important variables from MARS 1

```{r}
library(mgcv)

gam1 <- mgcv::gam(INS ~ s(ACCTAGE)+
                  s(DDABAL)+
                  s(DEP)+
                  s(CHECKS)+
                  s(TELLER)+
                  s(SAVBAL)+
                  s(ATMAMT)+
                  s(CDBAL)+
                  s(IRABAL)+
                  s(MMBAL)+
                  s(CCBAL)+
                  DDA+INV+CC+factor(PHONE_flag)+SAV, method = 'REML',family=binomial,data = final_df)

summary(gam1)

```

Did other group members use select = TRUE

Predicted probabilities

```{r}
test_gam1_df <- final_df

test_gam1_df$phat_gam1 <- predict(gam1, type = 'response')

roc_gam1 <- rocit(as.vector(test_gam1_df$phat_gam1), test_gam1_df$INS)

summary(roc_gam1)

```

AUC = 0.7977

MARS with cross validation

```{r}
mars2 <- earth(INS ~ ., data = final_df, glm = list(family=binomial), pmethod="cv", nfold=10,trace=3)
summary(mars2)

```

Importance: SAVBAL, CDBAL, DDA1, DDABAL, MMBAL, PHONE_flag, ACCTAGE, CHECKS, TELLER, ATMAMT, INV1, CC1, CCBAL, ..
(same as without CV)

Predicted probabilities

```{r}

test_mars2_df <- final_df

test_mars2_df$phat_mars2 <- predict(mars2, type = 'response')

roc_mars2 <- rocit(as.vector(test_mars2_df$phat_mars2), test_mars2_df$INS)

summary(roc_mars2)

```

AUC = 0.7995

GAM with all variables

```{r}
#data_types <- sapply(final_df, class)
#print(data_types)

#zero_columns <- sapply(final_df, function(col) all(col == 0))

# Print the column names
#print(names(final_df)[zero_columns])

library(mgcv)

gam2 <- mgcv::gam(INS ~ DDA + DIRDEP + NSF + SAV + ATM + CD + IRA + INV + MM + MMCRED + CC + CCPURC + SDB + INAREA + s(ACCTAGE) + s(DDABAL) + s(DEP) + s(DEPAMT) + s(CHECKS) + s(NSFAMT) + s(PHONE) + s(TELLER) + s(SAVBAL) + s(ATMAMT) + s(POS) + s(POSAMT) + s(CDBAL) + s(IRABAL) + s(INVBAL) + s(MMBAL) + s(CCBAL) + s(INCOME) + s(LORES) + s(HMVAL) + s(AGE) + s(CRSCORE) + ACCTAGE_flag + PHONE_flag + POS_flag + POSAMT_flag + INV_flag + INVBAL_flag + CC_flag + CCBAL_flag + CCPURC_flag + INCOME_flag + LORES_flag + HMVAL_flag + AGE_flag + CRSCORE_flag, method = 'REML',family=binomial,data = final_df)

summary(gam2)

```

Predicted probabilities

```{r}
test_gam2_df <- final_df

test_gam2_df$phat_gam2 <- predict(gam2, type = 'response')

roc_gam2 <- rocit(as.vector(test_gam2_df$phat_gam2), test_gam2_df$INS)

summary(roc_gam2)

```

AUC = 0.8006